Synthesis of Linear Functional Programs

### 2021-07-23 article/article.tex
Add example to end of evaluation, cut some text

The \synname\ synthesizer combines linear types with recursion,
parametric polymorphism, recursive algebraic data types, and
refinement types. The synthesizer is built on top of a system of proof-search for
linear logic. Proof search relates to program synthesis via the Curry-Howard
correspondence~\cite{}, which states that \emph{propositions in a logic} are
\emph{types}, and \emph{proofs} of those propositions are well-typed
\emph{programs} -- finding a proof of a proposition is finding a program with
that type.

@@

In intuitionistic sequent calculi, each connective has a so-called \emph{left}
and a \emph{right} rule, which effectively define how to decompose an ambient
assumption of a given proposition and how to prove a certain proposition is
true, respectively.  Andreoli's \emph{focusing} for linear logic~\cite{Andreolli
focusing} is a technique to remove non-essential nondeterminism from
proof-search by structuring the application of so-called invertible and
non-invertible inference rules. Andreoli observed that the connectives of linear
logic can be divided into two categories, dubbed synchronous and asynchronous.
Asynchronous connectives are those whose right rules are \emph{invertible}, i.e.
they can be applied eagerly during proof search without losing provability and
so the order in which these rules are applied is irrelevant, and whose left
rules are not invertible. Synchronous connectives are dual. The asynchronous
connectives are $\lolli$ and $\with$ and the synchronous are
$\tensor,\textbf{1},\oplus,\bang$.

@@

Putting together linear logic and linear lambda calculus through the Curry-Howard correspondence, resource
management, and focusing, we get the following core formal system
(inspired by~\cite{FPnotes...}) -- in which the
rule $\lolli$R is read: to synthesize a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and inversion context $\Omega$, synthesize a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x . M$ and the output linear context is
$\Delta'$.

@@

\mypara{Polymorphic Types} A polymorphic specification is a type of form
$\forall \overline{\alpha}.\ \tau$ where $\overline{\alpha}$ is a set of
variables that stand for any (non-polymorphic) type in $\tau$. Such a type is
also called a \emph{scheme}.  Synthesis of a scheme comprises of turning it into
a non-quantified type, and then treating its type variables uniformly.  First,
type variables are considered \emph{atomic types}, then, we instantiate the
bound variables of the scheme as described by the Hindley-Milner inference
method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names
for each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha \lolli
\alpha$ could be instantiated to $\alpha0 \lolli \alpha0$. We add a rule for
this, where $\forall \overline{\alpha}.\ \tau \sqsubseteq \tau'$ indicates type
$\tau'$ is an \emph{instantiation} of type scheme $\forall \overline{\alpha}.\
\tau$:
%
\begin{mathpar}
    \infer*[right=($\forall R$)]
    { \Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \tau' \Uparrow \and \forall
    \overline{\alpha}.\ \tau
    \sqsubseteq \tau'}
    {\Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \forall \overline{\alpha}.\
    \tau \Uparrow}
\end{mathpar}
%
As such, the construction of a derivation in which the only rule that can derive
an atom is the \textsc{init} rule corresponds to the synthesis of a program
where some expressions are treated agnostically (nothing constrains their type),
i.e.~a polymorphic program. The simplest example is the polymorphic function
\emph{id} of type $\forall \alpha .\ \alpha \lolli \alpha$. The program
synthesized from that specification is $\lambda x . x$, a lambda abstraction
that does not constrain the type of its parameter $x$ in any way.

@@

The desugaring module converts the frontend AST to a core AST without
syntatic sugar and using the locally-nameless
representation~\cite{locally nameless}, in which all bound variables
are represented with Debruijn indices rather than names (i.e.~using
the distance to the binding site in terms of the number of traversed
binders),
This way inference can be done without worrying about name conflicts.

@@

\mypara{Memoization} Some proofs attempt to synthesize the same
sub-goal with the same premises more than once. Since these proof
attempts might be expensive performance-wise, and we can rely on the
program's determinism (equal calls to synthesize result in equal
outcomes), we added memoization to our synthesizer -- an optimization
technique that stores function calls and returns cached values for
equal inputs.
%
For our synthesizer, two calls are equivalent when
contexts $\Theta, \Rho, \Gamma, \Delta$, $\Omega$, (or alternatively
the left focus), and the goal, are equivalent. Because comparing all
of these is slow, our implementation hashes the relevant values via a hashing
library (\emph{Hashable}~\cite{https://hackage.haskell.org/package/hashable}) and
uses the result as the key.
%
Furthermore, a branch we want to memoize
might fail, not providing any result -- we also want to record
synthesis failure, as it happens more often than successful
results. We capture failure of a LogicT computation using the $\mathit{ifte}$
function and represent failure or success information with the \texttt{Maybe}
type. We ran benchmarks to measure the impact of memoization on 
performance. It's quite noticeable in synthesis with multiple
functions in the unrestricted context, or more complex
specifications. For example, if we add a new function \emph{read} to
the unrestricted context in the \emph{array} problem seen in
\S~\ref{sec:overview}, synthesis with memoization will take
around $9s$, while synthesis without memoization doesn't terminate
after $15m$.
%
The implementation could be improved by better combining the
memoization with the LogicT monad transformer (remembering decisions across
backtracking branches) and by finetuning key-equality by ignoring context
components that aren't relevant to the synthesis call, to name some
examples. We leave such optimizations to future work.

\mypara{Debugging} The rules described in the formal
system (\S~\ref{sec:formal_system}) to handle infinite recursion might seem obvious in
retrospective, however, when faced with a program that doesn't terminate, what
went wrong is not so clear. While we were able to figure out a solution for the
first couple of "infinite examples", the following problems were much harder,
taking sometimes full days to solve. To make the debugging experience better, we
developed a tracing system that prints information whenever a rule is applied,
alongside the stack of rules applied up until that one. This improved our
debugging speed considerably, especially when put together with a correct
derivation done by hand, with which we could match what was happening against
what was supposed to happen.

@@

The benchmarks\footnote{Considering the full pipeline pass; 2017 MBP; 8GB RAM;
2.3Ghz Intel Core i5; with benchmarking library Criterion} are separated into groups: theorems of linear logic, recursive
ADTs, refinements. The first table displays successful synthesis -- for each
goal we analyze the mean time and standard deviation, the conditions used to
guide the output, and the components (other functions) used in the resulting
program.

% TODO Verificar nomes dos linear logic theoresm...
{\scriptsize
\begin{center}
    \begin{tabular}{ |c|c|c|c|c| }
        \hline
        Group & Goal & Avg. time $\pm\,\sigma$ & Keywords & Components \\
        \hline
        \multirow{4}{4em}{Linear Logic Theor.} & uncurry & $133\mu s\pm 4.9\mu s$ &&\\ 
        & distributivity & $179\mu s\pm 5.0\mu s$ && \\ 
        & call by name & $196\mu s\pm 4.6\mu s$ && \\ 
        & 0/1 & $294\mu s\pm 5.3\mu s$ && \\ 
        \hline
        \multirow{7}{4em}{List} & map & $288\mu s\pm 7.2\mu s$  && \\ 
        & append & $292\mu s\pm 7.0\mu s$ & & \\ 
        & foldl & $1.69ms\pm 5.3\mu s$ & \emph{choose 1} & \\ 
        & foldr & $704\mu s\pm 10\mu s$ && \\ 
        & concat & $505\mu s\pm 18\mu s$ && \\
        & uncons & $215\mu s\pm 15\mu s$ && \\
        & reverse & $17.4ms\pm 515\mu s$ & \emph{reverse [1,2] == [2,1]} & \\ 
        \hline
        \multirow{2}{4em}{Maybe} & $>>=$ & $194\mu s\pm 5.3\mu s$ && \\ 
        & maybe & $161\mu s\pm 4.8\mu s$&& \\
        \hline
        \multirow{7}{4em}{State} & runState & $190\mu s\pm 6.8\mu s$ && \\ 
        & $>>=$ & $979\mu s\pm 23\mu s$ && \\
        & $>>=$ & $\infty$ & \emph{using (runState)} & \\
        & get & $133\mu s\pm 3.8\mu s$ && \\
        & put & $146\mu s\pm 3.4\mu s$ && \\
        & modify & $219\mu s\pm 4.9\mu s$ && \\
        & evalState & $156\mu s\pm 4.0\mu s$ && \\
        % \multirow{5}{4em}{Binary Tree} & singleton & 1s & yes & yes \\ 
        % & insert & 1s & yes & unordered \\
        % & map & 1s & yes & yes \\
        % & append & 1s & yes & appends to rightmost leaf \\
        % & join node & 1s & yes & joins at rightmost leaf \\
        % \hline
        % \multirow{7}{4em}{Map} & singleton & 1s & yes & yes \\ 
        % & insert & 1s & yes & unordered \\
        % & insertWithKey & 1s & yes & ignores function \\
        % & union & 1s & yes & ignores keys \\
        % & mapAccum & 1s & no & ?? \\
        % & map & 1s & yes & yes \\
        % & mapKeys & 1s & yes & yes \\
        \hline
        \multirow{1}{4em}{Misc} & either & $197\mu s\pm 5.3\mu s$ && \\ 
        \hline
        \multirow{2}{4em}{Array} & & & \emph{depth 3} & \emph{freeze, foldl} \\
        & array & $80ms\pm 870\mu s$ & \emph{using (foldl),depth 3} & \emph{newMArray,write} \\
        \hline
    \end{tabular}
  \end{center}
}

>>>>

The \synname\ synthesizer combines linear types with recursion,
parametric polymorphism, recursive algebraic data types, and
refinement types. The synthesizer is built on top of a system of proof-search for
linear logic. Proof search relates to program synthesis via the Curry-Howard
correspondence~\cite{curry:34,howard:80,DBLP:journals/cacm/Wadler15}, which states that \emph{propositions in a logic} are
\emph{types}, and \emph{proofs} of those propositions are well-typed
\emph{programs} -- finding a proof of a proposition is finding a program with
that type.

@@

In intuitionistic sequent calculi, each connective has a so-called \emph{left}
and a \emph{right} rule, which effectively define how to decompose an ambient
assumption of a given proposition and how to prove a certain proposition is
true, respectively.  Andreoli's \emph{focusing} for linear logic~\cite{10.1093/logcom/2.3.297} 
is a technique to remove non-essential nondeterminism from
proof-search by structuring the application of so-called invertible and
non-invertible inference rules. Andreoli observed that the connectives of linear
logic can be divided into two categories, dubbed synchronous and asynchronous.
Asynchronous connectives are those whose right rules are \emph{invertible}, i.e.
they can be applied eagerly during proof search without losing provability and
so the order in which these rules are applied is irrelevant, and whose left
rules are not invertible. Synchronous connectives are dual. The asynchronous
connectives are $\lolli$ and $\with$ and the synchronous are
$\tensor,\textbf{1},\oplus,\bang$.

@@

Putting together linear logic and linear lambda calculus through the Curry-Howard correspondence, resource
management, and focusing, we get the following core formal system
(inspired by~\cite{DBLP:conf/cade/ChaudhuriP05,fpnotes}) -- in which the
rule $\lolli$R is read: to synthesize a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and inversion context $\Omega$, synthesize a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x . M$ and the output linear context is
$\Delta'$.

@@

\mypara{Polymorphic Types} A polymorphic specification is a type of form
$\forall \overline{\alpha}.\ \tau$ where $\overline{\alpha}$ is a set of
variables that stand for any (non-polymorphic) type in $\tau$. Such a type is
also called a \emph{scheme}.  Synthesis of a scheme comprises of turning it into
a non-quantified type, and then treating its type variables uniformly.  First,
type variables are considered \emph{atomic types}, then, we instantiate the
bound variables of the scheme as described by the Hindley-Milner inference
method's~\cite{DBLP:journals/jcss/Milner78,10.2307/1995158} instantiation rule (put simply, generate fresh names
for each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha \lolli
\alpha$ could be instantiated to $\alpha0 \lolli \alpha0$. We add a rule for
this, where $\forall \overline{\alpha}.\ \tau \sqsubseteq \tau'$ indicates type
$\tau'$ is an \emph{instantiation} of type scheme $\forall \overline{\alpha}.\
\tau$:
%
\begin{mathpar}
    \infer*[right=($\forall R$)]
    { \Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \tau' \Uparrow \and \forall
    \overline{\alpha}.\ \tau
    \sqsubseteq \tau'}
    {\Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \forall \overline{\alpha}.\
    \tau \Uparrow}
\end{mathpar}
%
As such, the construction of a derivation in which the only rule that can derive
an atom is the \textsc{init} rule corresponds to the synthesis of a program
where some expressions are treated agnostically (nothing constrains their type),
i.e.~a polymorphic program. The simplest example is the polymorphic function
\emph{id} of type $\forall \alpha .\ \alpha \lolli \alpha$. The program
synthesized from that specification is $\lambda x . x$, a lambda abstraction
that does not constrain the type of its parameter $x$ in any way.

@@

The desugaring module converts the frontend AST to a core AST without
syntatic sugar and using the locally-nameless
representation~\cite{DBLP:journals/jar/Chargueraud12}, in which all bound variables
are represented with Debruijn indices rather than names (i.e.~using
the distance to the binding site in terms of the number of traversed
binders),
This way inference can be done without worrying about name conflicts.

@@

\mypara{Memoization} Some proofs attempt to synthesize the same
sub-goal with the same premises more than once. Since these proof
attempts might be expensive performance-wise, and we can rely on the
program's determinism (equal calls to synthesize result in equal
outcomes), we added memoization to our synthesizer -- an optimization
technique that stores function calls and returns cached values for
equal inputs.
%
For our synthesizer, two calls are equivalent when
contexts $\Theta, \Rho, \Gamma, \Delta$, $\Omega$, (or alternatively
the left focus), and the goal, are equivalent. Because comparing all
of these is slow, our implementation hashes the relevant values via a hashing
library (\emph{Hashable}~\cite{hashable}) and
uses the result as the key.
%
Furthermore, a branch we want to memoize
might fail, not providing any result -- we also want to record
synthesis failure, as it happens more often than successful
results. We ran benchmarks to measure the impact of memoization on 
performance. It's quite noticeable in synthesis with multiple
functions in the unrestricted context, or more complex
specifications. For example, if we add a new function \emph{read} to
the unrestricted context in the \emph{array} problem seen in
\S~\ref{sec:overview}, synthesis with memoization will take
around $9s$, while synthesis without memoization doesn't terminate
after $15m$.
%
The implementation could be greatly improved, but we
% by better combining the
% memoization with the LogicT monad transformer (remembering decisions across
% backtracking branches) and by finetuning key-equality by ignoring context
% components that aren't relevant to the synthesis call, to name some
% examples. We 
leave such optimizations to future work.

\mypara{Debugging} The rules described in the formal
system (\S~\ref{sec:formal_system}) to handle infinite recursion might seem obvious in
retrospective, however, when faced with a program that doesn't terminate, what
went wrong is not so clear.
% While we were able to figure out a solution for the
% first couple of "infinite examples", the following problems were much harder,
% taking sometimes full days to solve.
To make the debugging experience better, we
developed a tracing system that prints information whenever a rule is applied,
alongside the stack of rules applied so far.
% This improved our
% debugging speed considerably, especially when put together with a correct
% derivation done by hand, with which we could match what was happening against
% what was supposed to happen.

@@

The benchmarks\footnote{ran on an Intel Core i5 machine with 8GB of RAM using the benchmarking library
Criterion, and considering the full pipeline pass, from parsing to synthesis} are separated into groups: theorems of linear logic, recursive
ADTs, refinements. The first table displays successful synthesis -- for each
goal we analyze the mean time and standard deviation, the conditions used to
guide the output, and the components (other functions) used in the resulting
program.

% TODO Verificar nomes dos linear logic theoresm...
{\scriptsize
\begin{center}
    \begin{tabular}{ |c|c|c|c|c| }
        \hline
        Group & Goal & Avg. time $\pm\,\sigma$ & Keywords & Components \\
        \hline
        \multirow{4}{4em}{Linear Logic Theor.} & uncurry & $133\mu s\pm 4.9\mu s$ &&\\ 
        & distributivity & $179\mu s\pm 5.0\mu s$ && \\ 
        & call by name & $196\mu s\pm 4.6\mu s$ && \\ 
        & 0/1 & $294\mu s\pm 5.3\mu s$ && \\ 
        \hline
        \multirow{7}{4em}{List} & map & $288\mu s\pm 7.2\mu s$  && \\ 
        & append & $292\mu s\pm 7.0\mu s$ & & \\ 
        & foldl & $1.69ms\pm 5.3\mu s$ & \emph{choose 1} & \\ 
        & foldr & $704\mu s\pm 10\mu s$ && \\ 
        & concat & $505\mu s\pm 18\mu s$ && \\
        & uncons & $215\mu s\pm 15\mu s$ && \\
        & reverse & $17.4ms\pm 515\mu s$ & \emph{reverse [1,2] == [2,1]} & \\ 
        \hline
        \multirow{2}{4em}{Maybe} & $>>=$ & $194\mu s\pm 5.3\mu s$ && \\ 
        & maybe & $161\mu s\pm 4.8\mu s$&& \\
        \hline
        \multirow{7}{4em}{State} & runState & $190\mu s\pm 6.8\mu s$ && \\ 
        & $>>=$ & $979\mu s\pm 23\mu s$ && \\
        & $>>=$ & $\infty$ & \emph{using (runState)} & \\
        & get & $133\mu s\pm 3.8\mu s$ && \\
        & put & $146\mu s\pm 3.4\mu s$ && \\
        & modify & $219\mu s\pm 4.9\mu s$ && \\
        & evalState & $156\mu s\pm 4.0\mu s$ && \\
        % \multirow{5}{4em}{Binary Tree} & singleton & 1s & yes & yes \\ 
        % & insert & 1s & yes & unordered \\
        % & map & 1s & yes & yes \\
        % & append & 1s & yes & appends to rightmost leaf \\
        % & join node & 1s & yes & joins at rightmost leaf \\
        % \hline
        % \multirow{7}{4em}{Map} & singleton & 1s & yes & yes \\ 
        % & insert & 1s & yes & unordered \\
        % & insertWithKey & 1s & yes & ignores function \\
        % & union & 1s & yes & ignores keys \\
        % & mapAccum & 1s & no & ?? \\
        % & map & 1s & yes & yes \\
        % & mapKeys & 1s & yes & yes \\
        \hline
        \multirow{1}{4em}{Misc} & either & $197\mu s\pm 5.3\mu s$ && \\ 
        \hline
        \multirow{2}{4em}{Array} & & & \emph{depth 3} & \emph{freeze, foldl} \\
        & array~\ref{sec:overview} & $80ms\pm 870\mu s$ & \emph{using (foldl),depth 3} & \emph{newMArray,write} \\
        \hline
    \end{tabular}
  \end{center}
}

To further showcase the expressiveness of our synthesizer, we present as
examples the input and output program for \emph{reverse}.
Note how we can guide the synthesis using the assertion mechanism to 
supply an example:
{\small
\begin{minted}{haskell}
synth reverse :: List a -o List a -o List a
  | assert (reverse (Cons (1, Cons (2, Nil))) Nil) == (Cons (2, Cons (1, Nil))); 
\end{minted}

\begin{minted}{haskell}
reverse b c = case b of
    Nil -> c
  | Cons e*f -> reverse f (Cons (e, c));
\end{minted}
}
